---
title: "R Programming - Week 3: Loop Functions & Debugging"
author: "Mike Talley"
date: "2/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview

We already learned about loop functions, but the for / if loops (while useful for programming) can be kind of bulky on the command line. There are a handful of more compact loop functions that can make things much simpler. Often times, these functions have the word "apply" in them, as they are applied to a set of objects.

## lapply
**Loop over a list and evaluate a function on each element**
*What comes out is ALWAYS a list* 

lapply takes three arguments: (1), a list x; (2) a function or the name of a function FUN; (3) other arguments via its ... argument. *If* x *is not a list, it WILL BE COERCED using* as.list. Because of this, lapply will ALWAYS return a list, regardless of the class of input. 

```{r}
x <- 1:4  # Vector of 1-4
lapply(x, runif, min = 0, max = 10)   # Returns random number between 1 & 10. Will have a list of 1:4 elements.
```

lapply and the others make heavy use of *anonymous* functions. This is a function that exists ONLY within the lapply function, and nowhere else. So when lappy is done, the function/action ceases to exist. 

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))   # List of two matrices
x     # View matrices
lapply(x, function(elt) elt[,1])    # Define and immediately call a function that returns the first column.

# This function elt() only exists inside of lapply, and cannot be called later outside of this command line. 
```

## sapply
**Same as lapply but will try to simplify the result**

sapply will try to simplify lapply results if it's possible. 

- If the results is a list where every element is length 1, then a **vector** is returned.

- If the results is a list where every element is a vector of the same length (>1), a **matrix** is returned.

- If it *can't* figure things out, a **list** is returned. 

## apply
**Apply is used to evaluate a function (often anonymous) over the margins of an array**

Most often, it is used to apply a function to the rows or columns of a matrix. It can be used with general arrays, i.e. taking the average of an array of matrices. Contrary to popular belief, it's not really any faster than writing a for loop, but it works in one line, so you have to type less.

***Less typing is always better, because good programmers are always lazy***

str(apply)
function (X, MARGIN, FUN, ...)

- X is an array
- MARGIN is an interger vector indicating which margins should be "retained"
- FUN is a function to be applied
- ... is arguments to be passed to FUN

```{r}
# Function to return the mean of each of the columns in a matrix

x <- matrix(rnorm(200), 20, 10)   # Create matrix
apply(x, 2, mean)                 # Call the matrix, get mean of each column. Returns a vector of length=10.

# The "2" above is tricky to understand at first. A matrix has two dimensions (rows x cols). By listing "2", we are saying we want to keep the 2nd dimension (cols). So, the function is removing the rows, and just returning the means of each column. Below, we will will keep the rows (1), and discard the cols (2), to get the sum of each of the 20 rows. 

apply(x, 1, sum)
```

### **col/row sums and means**

For sums and means of matrix dimensions, we have shortcuts. These will be much faster than their apply counterparts, although you won't notice unless using large matrices.

- rowSums = apply(x, 1, sum)

- rowMeans = apply(x, 1, mean)

- colSums = apply(x, 1, sum)

- colMeans = apply(x, 1, mean)

### Other Ways to Apply

Apply can also creat a matrix when running a function. Take the previous example, with 20 rows and 10 columns. Below, we will ask for the 25% and 75% quantiles of the rows. To display this info, it will; 1 - "remove" the columns, 2 - calculate quantiles for each row, 3 - Output the answers in a matrix, with two rows for quantiles, and 20 columns, which are for each row in the prior matrix.

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(0.25, 0.75))
```

We can make a 3D array (ten 2x2 matrices on top of each other), and then "remove" the third dimension (10), to return a mean of that dimension. Trippy stuff!

```{r}
a <- array(rnorm(2 * 2 * 10), c(2, 2, 10))    # 3D array
apply(a, c(1,2), mean)                        # apply(array, remove 3rd dim, what did we do with it)
rowMeans(a, dims =  2)                        # Same idea, collapsing it into a 2D object.
```

## tapply
**apply a function over subsets of a vector**

Used to apply a function over subsets of a vector. Not quite sure why it's called tapply, but let's roll with it.
str(tapply)
function (x, INDEX, FUN = NULL, ..., simplify = TRUE)
  - x is the vector / data to be looked at.
  
  - INDEX is a factor or a list of factors (aka, how to select or subset men/women in a list of patients)
  
  - FUN is a function to be applied
  
  - ... contains other arguments to be passed to FUN
  
  - simplify, should we simplify the result?

## mapply
**multivariate version of lapply**

The function written within mapply can work across multiple lists in parallel, rather than needing a for loop. The number of arguments in this function must be at least the number of lists being passed. Below are two ways to write a function, the long way, and a shorter way using mapply.
```{r}
## First method
list(rep(1,4), rep(2,3), rep(3,2), rep(4,1))

## Second method
mapply(rep, 1:4, 4:1)
```

Another longer example... Below is a function to return a random selection of numbers with a specified mean and sd. 
```{r}
noise <- function(n, mean, sd) {
  rnorm(n, mean, sd)
}

## It works with a simple number, like n = 5)
noise(5, 1, 2)

## It does NOT however, work with a vector numbers, like n = 1:5. It will not return a vector like we want. 
noise(1:5, 1:5, 2)

## It WILL work if we pass the function through the mapply function. 
mapply(noise, 1:5, 1:5, 2)

## Typing this mapply function out the long way would look like...
list(noise(1, 1, 2), noise(2, 2, 2),
     noise(3, 3, 2), noise(4, 4, 2),
     noise(5, 5, 2))
```

## Split
**an auxiliary function to split up data. Extra useful when combined with lapply**

Split takes a vector or other objects, and splits it into groups determined by a factor or list of factors. Unlike tapply, which can be applied over different subsets of a vector, another option is to use split() in conjunction with lapply() or sapply(). 

str(split)
function (x, f, drop = FALSE, ...)

- x is a vector (or list) or data frame.

- f is a factor (or coerced to one) or a list of factors.

- drop indicates whether empty factors levels should be dropped. 

***split always returns a list back***

A common method is split() followed by lapply(). You could theoretically used mapply for the same (but more succint) task, but either way works. 

```{r}
## Split the airquality dataset up by month.
s <- split(airquality, airquality$Month)

## Get the mean for each of three variables, based on month. Note the NA values.
lapply(s, function(x)
  colMeans(x[, c("Ozone", "Solar.R", "Wind")])
)

## sapply will combine all of this into an easier to read matrix. Note that NAs have been removed. 
sapply(s, function(x)
  colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm = TRUE)
)
```

# Debugging Tools

## Diagnosing the Problem

Questions to ask when you get an error / warning:

- What was your input? How did you call the function?

- What were you expecting? Output, messages, other results?

- What did you get?

- How does what you get differ from what you were expecting?

- Were your expectations correct in the first place?

- Can you reproduce the problem (exactly)?

## Debugging Tools in R

- **Traceback**: Prints out the function call stack after an error occurs.

- **debug**: Flags a function for "debug" mode, which allows you to step through execution of a function, one line at a time. 

- **browser**: Suspends the execturion of a funciton wherever it is called and puts the function in debug mode.

- **trace**: Allows you to insert debugging code into a function at specific places.

- **recover**: Allows you to modify the error behavior so that you can browse the functino call stack.

Less elegant options also include inserting print/cat statements throughout your function. 

***Debugging tools are not a substitute for thinking! Write good code FIRST***