---
title: "R_Programming_Week_4"
author: "Mike Talley"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# str(): Compactly display the internal structur of an R object

This is one of the most valuable function/commands in R. It will output a brief summary of the structure of whatever object is passed to it. 

```{r}
str(airquality)

s <- split(airquality, airquality$Month)
str(s)
```

# Simulation

Simulation is also a super useful ability of R, which is frequently used in statistics. 

## Generating Random Numbers

We can use a variety of functions to generate random number distribution as well:

- rnorm: generate random Normal variates with a given mean and standard deviation.

- dnorm: evaluate the Normal probability density (with a given mean/SD) at a point (or vector of points)

- pnorm: evaluate the cumulative distribution function for a Normal distribution.

- rpois: generate random Poisson variates with a given rate.

Probability distributinos also generally have four functions associated with them (you can see some of them in the above list).

- d for density

- r for random number generation

- p for cumulative distribution

- q for quantile function

Normal distributions have standard (preset) means, sd, and other attributes. Use ? to check these out.

```{r}
x <- rnorm(10)
x

x <- rnorm(10, 20, 2)
x

summary(x)
```

**Anytime you generate random numbers, it is very important to set the random number generator "seed", to ensure re reproducability.** 

This is based on the idea that random numbers from a computer are not *actually* random, but they appear to be so, and that's all that matters. This means we can reproduce this generation, and other people can reproduce your work. 

```{r}
# Setting seed
set.seed(1)
rnorm(5)

# Different 
rnorm(5)

# Reset the seed to get the original output
set.seed(1)
rnorm(5)
```

## Generating Random Numbers from a Linear Model

We can do the same thing as above, and simulate a model using random numbers. 

```{r}
set.seed(20)
x <- rnorm(100)
e <- rnorm(100, 0, 2)
y <- 0.5 + 2 * x + e
summary(y)
plot(x, y)
```

## Random Sampling

The sample() function will randomly draw from a set of (scalar) objects, allowing you sample from arbitrary distributions. 

```{r}
set.seed(1)
sample(1:10, 4)
sample(1:10, 4)
sample(letters, 5)
sample(1:10)   ## permutation
sample(1:10)
sample(1:10, replace = TRUE) ## Sample w/ replacement
```

# R Profiler

Profiling is a systematic way to examine how much time is spent in different parts of a program. It can be useful when trying to optimize code. Sometimes, you may not notice the issues when you run it, but when it's used 1000times in a row, it may become more obvious that it needs optimized. 

## Principles of Optimizations

- Design first, then optimize. 

- Premature optimization is the root of all evil. 

- Measure / collect data, don't guess.

## System.time

Using system.time(), you can input an arbitrary R expression, and it will tell you the time (in seconds) it takes to exxecute an expression. If there is an error, it will give the time at which the error occured. 
It will return an object of the class "proc_time". *user time* is time charged to the CPUs for this expression, while *elapsed time* is "wall clock" time. 

Usually, these two times are relatively close. Elapsed time could be *greater than* user time if the CPU spends a lot of time waiting around. It could be *smaller than* user time if your computer has a lot of cores/processors working in sync. You could also be using parallel processing via the 'parallel' package.

Rprof() function starts the profiler in R. Do not use the system.time() function with the Rprof(), as they are not designed to run together. The output from this isn't particularly useful however, instead, we will use the summaryRprof() function. 
